---
title: Importing raw (unprocessed) Affymetrix microarray data
teaching: 10
exercises: 5
source: Rmd
questions:
  - "Can you think of how you might read the CEL files in an order that is guaranteed to match the  desired order?"
objectives:
  - Be able to obtain supplemental data
  - Be able to explain and use the differences between GEO data types.
  - Understand the concept of the ExpressionSet class of objects.
keypoints: 
  - GEO data types have enough similarities to allow data access, but enough differences to require specific type-specific steps.
  - The ExpressionSet class of object contains slots for different information associated  with a microarray experiment. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../bin/chunk-options.R")
library(GEOquery)

.starting_image <- ".02-RData"
.closing_image <- ".02-supplemental-RData"
if(file.exists(.starting_image)) {
    load(.starting_image,verbose=TRUE)
}
```

We now have processed data for two series, [GSE33146][GSE33146] and [GSE66417][GEE66417]. We need to get the _unprocessed_ data to understand the processing.

## Getting the raw data for the series using GEOquery

Getting the raw data (`CEL` files for Affymetrix data) is distinct from getting the processed ata. The `CEL` files can be pretty large, and sometimes the download can fail, even with a good connection.

Nonetheless, if you want to do it, `getGEOSuppFiles()` will download all the
supplementary files for the GEO accession. Note that it doesn't *process* or even *parse* the files,
since there are many different types of supplementary files on GEO and R doesn't know the format ahead of time. It does, however, return the file paths.

```{r, eval=FALSE}
## If you have an awesome connection and a lot of time
filePaths <- getGEOSuppFiles('GSE33146')
filePaths <- getGEOSuppFiles('GSE66417')
```

In the interest of time, don't do that. Instead, use the files you have downloaded from LumiNUS and
unpacked into a directory named for each series. You can leave them compressed.

## Reading CEL data using the `oligo` package.

The `oligo` package provides functions for handling Affymetrix data,
including CEL file data. The function `oligo::read.celfiles()` does
the work. This function takes in a vector of filenames as the
argument. We can manually type the file names into a vector, using the
`c()` function. Alternatively, we can use `list.celfiles()` command
provided by `oligoClasses`, which is installed when we install `oligo`.  `list.celfiles()` will list all the files
ending with the *.cel* extension (CEL files) in a directory, (by
default the current working directory). Helpfully:

- the argument `listGzipped=TRUE` will find compressed CEL files.
- the argument `full.names=TRUE` will include the _directory_ name, so the results can be passed to `read.celfiles()`. 

Therefore, the following needs to be done: 

1. Use `oligoClasseslist.celfiles(GSE33146)` with the appropriate arguments to generate a vector containing the name of all the CEL files.
2. Use `read.celfiles()` to read in the CEL files.

> ## Try it! 
>
> Try to read the CEL files for both data sets into R using the information provided above. 
>
> > ## Solution
> >
> > ```{r, eval=TRUE,message=FALSE}
> > library(oligo)
> > library(oligoClasses)
> > gse33146_celdata <- read.celfiles(list.celfiles('GSE33146',full.names=TRUE,listGzipped=TRUE))
> > gse66417_celdata <- read.celfiles(list.celfiles('GSE66417',full.names=TRUE,listGzipped=TRUE))
> > ```
> >
> {: .solution}
{: .challenge}


Once you have the CEL file data, you can get a pseudo-image of the chip intensities.

```{r}
image(gse33146_celdata[,1])
```

Note that `read.celfiles` loaded data packages from Bioconductor containing information on the platform design for each chip type.

> ## The risk of using  `list.celfiles()`
>
> Using `list.celfiles()` to provide the files to `read.celfiles()` can be risky.
> `list.celfiles()` provides a list of files in lexicographic order. This is *probably* the same
> order as the files in your GSE, but can you be sure? Ultimately, you need to use the metadata
> from the series to ensure that the rows of the phenoData match the columns of the assayData
> That is the subject of our next episode.
{: .warning}

> ## `oligo` or `affy`?
>
> The `affy` package also has a `list.celfiles()` function (with a
> slightly different interface), and offers a `read.AffyBatch()`
> function to read celfiles. However, the `affy` package can only work
> with 3′ biased Affymetrix arrays, so `oligo` is preferred
{: .warning}

```{r closeout, eval=TRUE, include=FALSE}
save(list=ls(),file=.closing_image)
Sys.sleep(1)
```
	
{% include site-links.md %} 
{% include links.md %} 
