---
title: Working with experimental metadata
teaching: 10
exercises: 5
source: Rmd
objectives:
- Be able to use metadata from GEO objects to construct useful R data objects
- Be able to use read.celfiles in combination with your own pData object to ensure data integrity
questions:
- "If an experiment were more complicated, and included several independent variables, would you be able to create a phenoData object from the GSE?"
keypoints: 
- GEO metadata can be cast into R data objects for analysis. The details are up to the user.
- Using proper phenoData to describe an experiment helps to ensure reproducibility and avoid reading in files out of order
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../bin/chunk-options.R")
library(oligo)
library(GEOquery)
.starting_image <- ".02-supplemental-RData"
.closing_image <- ".03-RData"
if(file.exists(.starting_image)) {
    load(.starting_image,verbose=TRUE)
    }

if (class(gse33146) == "list") {
    gse33146 <- gse33146[[1]]
}
```

We have successfully read CEL files into an R object, but we
haven't provided any information about the experimental design, which would
normally be provided in a phenoData object.  In the case of our two
experiments, the experimental designs are pretty simple. Let's see if
we can find the information we need from the metadata.

## What you should have

If you have been following this tutorial, you should have the following objects:

- An _ExpressionSet_ object for GSE33146 (the processed data). Let's assume this is called `gse33146`.
- An _ExpressionSet_ object for GSE66417  (the processed data). Let's assume this is called `gse66417`.
- An _ExpressionFeatureSet_ object for GSE33146 (the raw data). Let's assume this is called `gse33146_celdata`. 
- A _GeneFeatureSet_ object for GSE66417 (the raw data). Let's assume this is called `gse66417_celdata`. 
 
> ## Why do the raw data for the two experiments appear to be different classes? 
>
> The GSE33146 raw data is of class ExpressionFeatureSet, but the
> GSE66417 data is of class GeneFeatureSet. Raw Affymetrix data is all
> some form of "FeatureSet" data, but different array designs are
> _extensions_ of the abstract FeatureSet class. The first series is
> from a 3′-biased array design, and the second is from a gene-based
> design. The classes represent those different design types.
{: .callout}

## GSE33146

In the simple case, we already have an ExpressionSet object, and we can use the pData from it. 

```{r tidycleaning, message=FALSE}
library(tidyverse)
varLabels(gse33146)
# It looks like the variables of interest are "cell line:ch1" and "treatment:ch1", and the sample accessions are in "geo_accession"
gse33146$supplementary_file
pd <- pData(gse33146)
pd['cel_file'] <- str_split(pd$supplementary_file,"/") %>% map_chr(tail,1)
```

With this simple trick, we now can re-read out CEL files, and _guarantee_ that they are in the correct order as the experimental data we have from getGEO.

```{r, message=FALSE}
gse33146_celdata <- read.celfiles(paste0('GSE33146/',pd$cel_file),phenoData=phenoData(gse33146))
```
We get a warning because some extra information about detection channels is added, but no matter: we have already gotten the treatment conditions attached to our data!

```{r}
pData(gse33146_celdata)[,c("geo_accession","cell line:ch1","culture medium:ch1")]
```

## GSE66417

We can do the same thing with our second data set, but we can see that the experimental  layout is more complicated

```{r tidycleaning2, message=FALSE}
library(tidyverse)
varLabels(gse66417)
# It looks like the variables of interest are "cell type:ch1" and "treatment:ch1", and the sample accessions are in "geo_accession"
pd <- pData(gse66417)
pd['cel_file'] <- str_split(pd$supplementary_file,"/") %>% map_chr(tail,1)
```

We can repeat our simple trick to guarantee our cel files and our metadata are aligned.

```{r, message=FALSE}
gse66417_celdata <- read.celfiles(paste0('GSE66417/',pd$cel_file),phenoData=phenoData(gse66417))
```

```{r}
pData(gse66417_celdata)[,c("geo_accession","cell type:ch1","treatment:ch1")]
```
In this experiment, two coviariates are modified: cell type and treatment!

```{r closeout, eval=TRUE,include=FALSE}
save(list=ls(),file=.closing_image)
Sys.sleep(1)
```

{% include site-links.md %} 
{% include links.md %} 
